---
title: "Research 3/8/2022"
author: "Grace Rade"
date: "3/8/2022"
output: html_document
---
```{r}
library(tidyverse)
library(rio)
```

# Load datasets 
```{r}
mort_1979 <- import("CompressedMortality1979_1998_22.txt")
mort_1999 <- import("CompressedMortality1999_2016_2copy2.txt")

mort_hispanic <- read.delim("Compressed Mortality, 1999-2016copy.txt") %>% 
  filter(Hispanic.Origin == "Hispanic or Latino") %>% 
  rename(Race = Hispanic.Origin, Race.Code = Hispanic.Origin.Code)
mort_1979_race <-  read.delim("CompressedMortality1979-1998.txt")
mort_1999_race <- read.delim("CompressedMortality_1999_2016.txt")

mort_hispanic$Population <- as.numeric(mort_hispanic$Population)
```


# Join the full state population datasets together
```{r}
total_state_mort <- mort_1979 %>% 
  full_join(mort_1999, by = c("State", "Year", "Population", "Deaths")) %>% 
  select(State, Year, Population) %>% 
  rename(State_Population = Population)
```

# Join the by race and gender data sets together
```{r}
mort_race <- mort_1979_race %>%
  full_join(mort_1999_race, by = c("State", "Year", "Population", "Deaths", "Race", "Gender.Code"))

mort_race <- mort_race %>% 
  full_join(mort_hispanic, by = c("State", "Year", "Population", "Deaths", "Race", "Gender.Code")) %>% 
  select(State, Year, Gender.Code, Race, Population, Deaths) %>% 
  rename(Group_Population = Population)
```

#Join the state population and race/gender population data frames together
```{r}
mort_state_race <- mort_race %>% 
  full_join(total_state_mort, by = c("State", "Year")) %>%
  filter(Race == "Black or African American" | Race == "Hispanic or Latino" | Race == "White")
```

#Divide the population into subgroups and calculate population proportion, crude mortality rate, weighted mortality rate

#Black Mortality
```{r}
black_male_mort <- mort_state_race %>% 
  filter(Race == "Black or African American", Gender.Code == "M") %>% 
  group_by(State, Year) %>% 
  summarise(population_proportion = (Group_Population/State_Population), black_male_mortality = (Deaths/State_Population * 100000), black_male_mortality_weighted = black_male_mortality/population_proportion)
```

```{r}
black_female_mort <- mort_state_race %>% 
  filter(Race == "Black or African American", Gender.Code == "F") %>% 
  group_by(State, Year) %>% 
  summarise(population_proportion = (Group_Population/State_Population), black_female_mortality = (Deaths/State_Population * 100000), black_female_mortality_weighted = black_female_mortality/population_proportion)
```

# Hispanic mortality
```{r}
hispanic_male_mort <- mort_state_race %>% 
  filter(Race == "Hispanic or Latino", Gender.Code == "M") %>% group_by(State, Year) %>% 
  summarise(population_proportion = (Group_Population/State_Population), hispanic_male_mortality = (Deaths/State_Population * 100000), hispanic_male_mortality_weighted = hispanic_male_mortality/population_proportion)
```

```{r}
hispanic_female_mort <- mort_state_race %>% 
  filter(Race == "Hispanic or Latino", Gender.Code == "F") %>% group_by(State, Year) %>% 
  summarise(population_proportion = (Group_Population/State_Population), hispanic_female_mortality = (Deaths/State_Population * 100000), hispanic_female_mortality_weighted = hispanic_female_mortality/population_proportion)
```

# White mortality
```{r}
white_male_mort <- mort_state_race %>% 
  filter(Race == "White", Gender.Code == "M") %>% 
  group_by(State, Year) %>%
  summarise(population_proportion = (Group_Population/State_Population), white_male_mortality = (Deaths/State_Population * 100000), white_male_mortality_weighted = white_male_mortality/population_proportion)
```

```{r}
white_female_mort <- mort_state_race %>% 
  filter(Race == "White", Gender.Code == "F") %>% 
  group_by(State, Year) %>%
  summarise(population_proportion = (Group_Population/State_Population), white_female_mortality = (Deaths/State_Population * 100000), white_female_mortality_weighted = white_female_mortality/population_proportion)
```

# Join all the data frames together for a master dataset 
```{r}
white_mort <- white_female_mort %>% 
  full_join(white_male_mort, by = c("Year", "State")) %>% 
  select(-population_proportion.x, -population_proportion.y)
```

```{r}
black_mort <- black_male_mort %>% 
  full_join(black_female_mort, by = c("Year", "State")) %>% 
  select(-population_proportion.x, -population_proportion.y)
```

```{r}
hispanic_mort <- hispanic_male_mort %>% 
  full_join(hispanic_female_mort, c("Year", "State")) %>% 
  select(-population_proportion.x, -population_proportion.y)
```

```{r}
full_mort_weighted <- black_mort %>% 
  full_join(hispanic_mort, by = c("State", "Year"))

full_mort_weighted <- full_mort_weighted %>% 
  full_join(white_mort, by = c("State", "Year"))
```

```{r}
export(full_mort_weighted, "full_mort_weighted.csv")
```

